Supporting Information:\\Conformational plasticity of phospholipid headgroups in simulations and experiments

### 2021-04-27 manuscriptPGPEsuppl.tex
Update pds

\clearpage
\section{Evaluation of simulations against NMR experiments}
\subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}

The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}. Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments. Most importantly for this work, the CHARMM36 captures the distinct headgroup order parameters for PG and PS lipids observed in NMR experiments (Figs. \ref{HGorderParameters} and \ref{structures} in the main text).

It should be noted that the PG headgroup is biologically abundant R enantiomer in all simulations, while our $^{13}$C NMR experiments has a racemic mixture. Nevertheless, previous $^{2}$H NMR experiments comparing results between different enantiomers concluded that the structural differences between these are minor \cite{wohlgemuth80}.

@@

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}. A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935]. A symmetric lipid distribution was used. The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid. The simulated system was minimized, equilibrated and production run. Production simulation was 500-ns-long, which the last 100 ns were analysed.
For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs. For water, the SETTLE method was applied. The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps. The pressure was kept constant with a semi-isotropic scheme. The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1. Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh. van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm. Long-range dispersion correction to the pressure and energy was added.

@@

%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.} A pure POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935]. A symmetric lipid distribution was used. The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid. The simulated system was minimized, equilibrated and production run. The production simulation was 500-ns-long, which the last 100 ns were analysed.
For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs. For water, the SETTLE method was applied. The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps. The pressure was kept constant with a semi-isotropic scheme. The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1. Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh. van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm. Long-range dispersion correction to the pressure and energy was added.

>>>>

\clearpage
\section{Evaluation of simulations against NMR experiments}
\subsection{Conformational ensembles of headgroup and glycerol backbone in PE and PG lipids}
The quality of PE and PG headgroup conformational ensembles in different simulations are evaluated against NMR experiments in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}. Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments. Most importantly for this work, the CHARMM36 captures the distinct headgroup order parameters for PG and PS lipids observed in NMR experiments (Figs. \ref{HGorderParameters} and \ref{structures} in the main text).
It should be noted that the PG headgroup is biologically abundant R enantiomer in all simulations, while our $^{13}$C NMR experiments has a racemic mixture. Nevertheless, previous $^{2}$H NMR experiments comparing results between different enantiomers concluded that the structural differences between these are minor \cite{wohlgemuth80}.

@@

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}.
%A pure POPE system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935]. A symmetric lipid distribution was used. The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid. The simulated system was minimized, equilibrated and production run. Production simulation was 500-ns-long, which the last 100 ns were analysed.
%For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs. For water, the SETTLE method was applied. The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps. The pressure was kept constant with a semi-isotropic scheme. The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1. Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh. van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm. Long-range dispersion correction to the pressure and energy was added.

@@

\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.} A pure POPG system was built with CHARMM-GUI Membrane builder [https://doi.org/10.1021/acs.jctc.5b00935]. A symmetric lipid distribution was used. The bilayer contains 500 lipids and 50 CHARMM TIP3P Water model per lipid. The simulated system was minimized, equilibrated and production run. The production simulation was 500-ns-long, which the last 100 ns were analysed.
For production simulation, a Leap-Frog algorithm was applied for integrating Newton’s equations of motion, and the LINCS algorithm was used to constrain all covalent bonds, allowing a time step of 2 fs. For water, the SETTLE method was applied. The simulation was performed in the NPT ensemble where the temperature was kept constant with the V-rescale thermostat (JCP 126, 014101, https://doi.org/10.1063/1.2408420) with water together with eventual ions and lipids coupled to separate thermostats with a coupling constant of 0.1 ps. The pressure was kept constant with a semi-isotropic scheme. The Parrinello–Rahman barostat was used to keep the pressure at 1 bar with a coupling constant of 1 ps and a compressibility of 4.6 × 10–5 bar–1. Long-range electrostatic interactions with a real-space cutoff of 1.0 nm were treated by a particle mesh Ewald scheme (PME) with a Fourier spacing of 0.12 nm and a fourth-order interpolation to the Ewald mesh. van der Waals interactions were treated with a Lennard-Jones (LJ) potential with a cutoff of 1.0 nm. Long-range dispersion correction to the pressure and energy was added.
